---
title: "WNV Africa-Europe"
author: "Andrea Silverj"
date: '2023-06-29'
output: html_document
---

# -----[Data analysis and visualisation]-----

Set working directory and load libraries:
```{r message=FALSE, warning=FALSE}
setwd("C:/Users/silverja/Projects/wnv_EU-Africa/plots/first_revision")
library("ggplot2")
library("tidyverse")
library("ggtree")
library("ips")
library("treeio")
library("tidytree")
library("ggimage")
library("ggnewscale")
library("phytools")
library("phangorn")
library("writexl")
library("readxl")
library("extrafont")
library("TDbook")
library("grid")
library("gridExtra")
library("cowplot")

```

#########
# Trees #

# Figure 1b - Bayesian molecular clock of WNV L1
```{r}
# Load tree file and save information as a data frame
wnvtreeL1_bay <- read.beast("clock_wnv_L1_Afr-Eur_trimmed_tips_location_gamma4_relaxed_logN_bsky_1b.tre")
wnvtreeL1_bay_tib <- as_tibble(wnvtreeL1_bay)
wnvtreeL1_bay_df <- as.data.frame(wnvtreeL1_bay_tib)

# Extract infos from header
wnvtreeL1_bay_df_annp1 <- wnvtreeL1_bay_df %>% separate(label, into=c("id","place","date"), sep="[|]", remove=FALSE)

# Add node shapes and colours based on posterior supports
wnvtreeL1_bay_df_ann_toexport <- wnvtreeL1_bay_df_annp1 %>% mutate(posteriorsupport=ifelse(posterior >= 0.9, "high", "low"))

# Set colours
colour_region=c(low="#E69F00", high="transparent", africa="#D55E00", europe="#0072B2", middleeast="black")

# Save this data set and do manual curation #
# write_xlsx(wnvtreeL1_bay_df_ann_toexport, path="wnvtreeL1_bay_df_ann_exp.xlsx")

# Import the taxonomy curated table #
wnvtreeL1_bay_df_ann <- read_excel("wnvtreeL1_bay_df_ann_exp.xlsx")

# Plot the tree
wnvL1_tree_plot <- ggtree(wnvtreeL1_bay, size=1.1,  mrsd="2022-10-03")+
  aes(color=wnvtreeL1_bay_df_ann$continent)+
  theme_tree2(axis.line.x = element_line(size=1.1), 
  axis.ticks.x = element_line(size=1.1), axis.text.x = element_text(size=9))+
  theme(legend.position="top")+
  ggtitle("WNV-L1 timetree")+
  
  geom_range(range='height_0.95_HPD', color='#1820E4', alpha=.5, size=1.2, linetype=1, center="height_median")+
  geom_nodepoint(aes(color=wnvtreeL1_bay_df_ann$posteriorsupport), size=2.2)+scale_x_continuous(breaks = c(1850,1860,1870,1880,1890,1900,1910,1920,1930,1940,1950,1960,1970,1980,1990,2000,2010,2020))+
  geom_tiplab(aes(label=wnvtreeL1_bay_df_ann$label_study), size=2.5, offset=0.5)+
  labs(color="Posterior:")+
  scale_color_manual(values=colour_region, na.value="#000000")

# Save
ggsave("clock_L1-Afr-Eur_cc.pdf", width = 25, height = 28, units = "cm", limitsize = FALSE)

```

# Figure 1a - Phylogeography legend
```{r}
# Plot the legend to an empty plot, using data from the Spread3 plot
# Areas
legwnv1areasP <- wnvtreeL1_bay_df_ann %>% ggplot(aes(location1, location2)) + 
  geom_point(aes(colour = height_median), size = 3) +
  scale_color_gradient(high = "white", low = "black", limits = c(77.2,0.3), trans = 'reverse')+
    theme(legend.position="top")
legwnv1areas <- get_legend(legwnv1areasP)
remove(legwnv1areasP)
grid.newpage()                              
grid.draw(legwnv1areas)
ggsave("clock_L1-Afr-Eur_cc_legend_areas.pdf", width = 25, height = 28, units = "cm", limitsize = FALSE)

# Points
legwnv1pointsP <- wnvtreeL1_bay_df_ann %>% ggplot(aes(location1, location2)) + 
  geom_point(aes(colour = height_median), size = 3) +
  scale_color_gradient(low = "#A50026", high = "white", limits = c(103.7, 0.1), trans = 'reverse')+
    theme(legend.position="top")
legwnv1points <- get_legend(legwnv1pointsP)
grid.newpage()                              
grid.draw(legwnv1points)
ggsave("clock_L1-Afr-Eur_cc_legend_points.pdf", width = 25, height = 28, units = "cm", limitsize = FALSE)

```

# Figure 2b - Bayesian molecular clock of WNV L2
```{r}
# Load tree file and save information as a data frame
wnvtreeL2_bay <- read.beast("clock_wnv_L2_Afr-Eur_trimmed_tips_location_gamma4_relaxed_logN_bsky_1b.tre")
wnvtreeL2_bay_tib <- as_tibble(wnvtreeL2_bay)
wnvtreeL2_bay_df <- as.data.frame(wnvtreeL2_bay_tib)

# Extract infos from header
wnvtreeL2_bay_df_annp1 <- wnvtreeL2_bay_df %>% separate(label, into=c("id","place","date"), sep="[|]", remove=FALSE)

# Add node shapes and colours based on posterior supports
wnvtreeL2_bay_df_ann_toexport <- wnvtreeL2_bay_df_annp1 %>% mutate(posteriorsupport=ifelse(posterior >= 0.9, "high", "low"))

# Set colours
colour_region=c(low="#E69F00", high="transparent", africa="#D55E00", europe="#0072B2", middleeast="black")

# Save this data set and do manual curation #
# write_xlsx(wnvtreeL2_bay_df_ann_toexport, path="wnvtreeL2_bay_df_ann_exp.xlsx")

# Import the taxonomy curated table #
wnvtreeL2_bay_df_ann <- read_excel("wnvtreeL2_bay_df_ann_exp.xlsx")

# Plot the tree
wnvL2_tree_plot <- ggtree(wnvtreeL2_bay, size=1.1,  mrsd="2022-11-18")+
  aes(color=wnvtreeL2_bay_df_ann$continent)+
  theme_tree2(axis.line.x = element_line(size=1.1), 
  axis.ticks.x = element_line(size=1.1), axis.text.x = element_text(size=9))+
  theme(legend.position="top")+
  ggtitle("WNV-L2 timetree")+
  geom_range(range='height_0.95_HPD', color='#1820E4', alpha=.5, size=1.2, linetype=1, center="height_median")+
  geom_nodepoint(aes(color=wnvtreeL2_bay_df_ann$posteriorsupport), size=2.2)+
  scale_x_continuous(breaks = c(1590,1850,1860,1870,1880,1890,1900,1910,1920,1930,1940,1950,1960,1970,1980,1990,2000,2010,2020))+
  geom_tiplab(aes(label=wnvtreeL2_bay_df_ann$label_study), size=2.5, offset=0.5)+
  labs(color="Posterior:")+
  scale_color_manual(values=colour_region, na.value="#000000")

# Save
ggsave("clock_L2-Afr-Eur_cc.pdf", width = 40, height = 25, units = "cm", limitsize = FALSE)

```

# Figure 2a - Phylogeography legend L2
```{r}
# Plot the legend to an empty plot, using data from the Spread3 plot
# Areas
legwnv1areasP <- wnvtreeL2_bay_df_ann %>% ggplot(aes(location1, location2)) + 
  geom_point(aes(colour = height_median), size = 3) +
  scale_color_gradient(high = "white", low = "black", limits = c(169.9,2.1), trans = 'reverse')+
  theme(legend.position="top")
legwnv1areas <- get_legend(legwnv1areasP)
remove(legwnv1areasP)
grid.newpage()                              
grid.draw(legwnv1areas)
ggsave("clock_L2-Afr-Eur_cc_legend_areas.pdf", width = 25, height = 28, units = "cm", limitsize = FALSE)

# Points
legwnv1pointsP <- wnvtreeL2_bay_df_ann %>% ggplot(aes(location1, location2))+
  geom_point(aes(colour = height_median), size = 3) +
  scale_color_gradient(low = "#A50026", high = "white", limits = c(289.2, 0.0), trans = 'reverse')+
  theme(legend.position="top")
legwnv1points <- get_legend(legwnv1pointsP)
grid.newpage()                              
grid.draw(legwnv1points)
ggsave("clock_L2-Afr-Eur_cc_legend_points.pdf", width = 25, height = 28, units = "cm", limitsize = FALSE)

```

# Supplementary Figure 3 - Sequence downsampling L1
```{r}
# Load and root the tree 
wnv1_ml_tree <- read.tree("RAxML_bipartitions.wnvL1_afr-eur_noreco_trimmed.tre")
nseqs <- as.integer(length(wnv1_ml_tree$tip.label))
wnv1_ml_tree_r <- midpoint(wnv1_ml_tree)

# Convert data stored in the tree into a dataframe and save all in a file
wnv1_ml_tree_r_tib <- as_tibble(wnv1_ml_tree_r)
wnv1_ml_tree_r_df <- as.data.frame(wnv1_ml_tree_r_tib)

bvalues <- wnv1_ml_tree_r_df$label[(nseqs+2):nrow(wnv1_ml_tree_r_df)]

wnv1_ml_tree_r_df$bootstrap_support <- NA

wnv1_ml_tree_r_df$bootstrap_support[(nseqs+2):nrow(wnv1_ml_tree_r_df)] <- bvalues

wnv1_ml_tree_r_df$label[(nseqs+2):nrow(wnv1_ml_tree_r_df)] <- NA

wnv1_ml_tree_r_df_annp1 <- wnv1_ml_tree_r_df %>% separate(label, into=c("id","place","date"), sep="[|]", remove=FALSE)

wnv1_ml_tree_r_df_annp2 <- wnv1_ml_tree_r_df_annp1 %>% mutate(bootstrap_category=ifelse(as.integer(bootstrap_support) >= 90, "high", "low"))

id_selected_seqsL1 <- drop_na(data.frame(label=wnvtreeL1_bay_df$label, selected="yes"))

wnv1_ml_tree_r_df_ann <- full_join(wnv1_ml_tree_r_df_annp2, id_selected_seqsL1)

# Set colours
colour_region=c(high="#E69F00", low="transparent", yes="darkgreen", "NA"="#000000")

# Annotate clusters
american_cluster <- data.frame(node=c(250), type=c("Clade"))

# Plot and save the tree
wnv1_ml_tree_r_gg <- ggtree(wnv1_ml_tree_r)+
  geom_tiplab(aes(color=wnv1_ml_tree_r_df_ann$selected), size=2.5)+
  geom_nodepoint(aes(color=wnv1_ml_tree_r_df_annp2$bootstrap_category))+
  geom_hilight(data=american_cluster, aes(node=node),type = "roundrect", alpha=0, colour="#373332", extend=0.022)+
  geom_cladelab(node=250, label="American cluster", colour="black", horizontal=TRUE, barsize=1, angle=90, hjust=0.5, offset=0.02, offset.text=0.001)+
  aes(color=wnv1_ml_tree_r_df_ann$selected)+
  scale_color_manual(values=colour_region, na.value="#000000")
 
ggsave("Supplementary_Figure_3_Rev1.pdf", width = 30, height = 60, units = "cm", limitsize = FALSE)

```

# Supplementary Figure 4 - Sequence downsampling L2
```{r}
# Load and root the tree 
wnv2_ml_tree <- read.tree("RAxML_bipartitions.wnvL2_afr-eur_noreco_trimmed.tre")
nseqs <- as.integer(length(wnv2_ml_tree$tip.label))
wnv2_ml_tree_r <- midpoint(wnv2_ml_tree)

# Convert data stored in the tree into a dataframe and save all in a file
wnv2_ml_tree_r_tib <- as_tibble(wnv2_ml_tree_r)
wnv2_ml_tree_r_df <- as.data.frame(wnv2_ml_tree_r_tib)

bvalues <- wnv2_ml_tree_r_df$label[(nseqs+2):nrow(wnv2_ml_tree_r_df)]

wnv2_ml_tree_r_df$bootstrap_support <- NA

wnv2_ml_tree_r_df$bootstrap_support[(nseqs+2):nrow(wnv2_ml_tree_r_df)] <- bvalues

wnv2_ml_tree_r_df$label[(nseqs+2):nrow(wnv2_ml_tree_r_df)] <- NA

wnv2_ml_tree_r_df_annp1 <- wnv2_ml_tree_r_df %>% separate(label, into=c("id","place","date"), sep="[|]", remove=FALSE)

wnv2_ml_tree_r_df_annp2 <- wnv2_ml_tree_r_df_annp1 %>% mutate(bootstrap_category=ifelse(as.integer(bootstrap_support) >= 90, "high", "low"))

id_selected_seqsL2 <- drop_na(data.frame(label=wnvtreeL2_bay_df$label, selected="yes"))

wnv2_ml_tree_r_df_ann <- full_join(wnv2_ml_tree_r_df_annp2, id_selected_seqsL2)

# Set colours
colour_region=c(high="#E69F00", low="transparent", yes="darkgreen", "NA"="#000000")

# Annotate clusters
american_cluster <- data.frame(node=c(250), type=c("Clade"))

# Plot and save the tree
wnv2_ml_tree_r_gg <- ggtree(wnv2_ml_tree_r)+
  geom_tiplab(aes(color=wnv2_ml_tree_r_df_ann$selected), size=2.5)+
  geom_nodepoint(aes(color=wnv2_ml_tree_r_df_annp2$bootstrap_category))+
  aes(color=wnv2_ml_tree_r_df_ann$selected)+
  scale_color_manual(values=colour_region, na.value="#000000")
 
ggsave("Supplementary_Figure_4_Rev1.pdf", width = 30, height = 70, units = "cm", limitsize = FALSE)

```

# Supplementary Figure 7 - Sensitivity analysis: effect of sampling on the location of inferred ancestral nodes
```{r}
# Load tree file and save information as a data frame
wnvtree_bay <- read.beast("clock_L1-Afr-Eur_ss_50_trimmed_tips_location_gamma4_relaxed_logN_bsky_250m.tre")
wnvtree_bay_tib <- as_tibble(wnvtree_bay)
wnvtree_bay_df <- as.data.frame(wnvtree_bay_tib)

# Extract infos from header
wnvtree_bay_df_annp1 <- wnvtree_bay_df %>% separate(label, into=c("id","place","date"), sep="[|]", remove=FALSE)

# Add node shapes and colours based on posterior supports
wnvtree_bay_df_ann_toexport <- wnvtree_bay_df_annp1 %>% mutate(posteriorsupport=ifelse(posterior >= 0.9, "high", "low"))

# Save this data set and do manual curation #
write_xlsx(wnvtree_bay_df_ann_toexport, path="clock_L1-Afr-Eur_ss_50_trimmed_tips_location_gamma4_relaxed_logN_bsky_250m.xlsx")

# Set colours and shapes
colour_analysis=c(full="black", sen75="#44AA99", sen50="#DDCC77", sen30="#CC6677")

shape_analysis=c(full=16, sen75=3, sen50=15, sen30=17)

# Scatterplot - Location1 Vs Location2
# L1
location_scatter_L1_df <- read_excel("wnvSensitivity_location.xlsx", sheet="sensitivityL1")

location_scatter_L1_plot <- ggplot(location_scatter_L1_df, aes(location2, location1, color=analysis, shape=analysis))+ geom_point() + xlab("Latitude") + ylab("Longitude") + scale_color_manual(values=colour_analysis, na.value="#000000") + scale_shape_manual(values=shape_analysis, na.value="#000000") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))

# L2
location_scatter_L2_df <- read_excel("wnvSensitivity_location.xlsx", sheet="sensitivityL2")

location_scatter_L2_plot <- ggplot(location_scatter_L2_df, aes(location2, location1, color=analysis, shape=analysis))+  geom_point() + xlab("Latitude") + ylab("Longitude") + theme_bw() + scale_color_manual(values=colour_analysis, na.value="#000000") + scale_shape_manual(values=shape_analysis, na.value="#000000") + theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))

```
